\vspace{-2.0em} Move Schedules: Fast persistence computations in coarse dynamic settings \footnote{ \textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758. }

### 2023-10-13 move_schedules.tex
Finalizing the next revision prior to final couple checks for grammar and typos

\begin{figure}[!htb]
    \centering
    \includegraphics[height=1.25in]{circle_complex.png}
    \caption{Top: A video of an expanding annulus. Bottom: Sublevel-set filtrations, via pixel intensity, of a Freudenthal triangulation of the plane.}
    \label{fig:vidExample}
\end{figure}

@@

%\begin{remark}
%	Conceptually, there is a conceptual similarity between the cyclic permutation that moves a simplex in a filtration the \emph{cycle sort}~\cite{haddon1990cycle}, a simple sorting algorithm method that has $O(n^2)$ time complexity for general inputs and $O(n)$ complexity for permutations.
%	Indeed, just as cycle sort rotates elements within a \emph{cycle} (in the permutation sense) using $O(1)$ auxiliary storage, the move right operation  
%\end{remark}
Regarding the complexity of move operations, which clearly depend on the sparsity of $R$ and $V$, we recall the proposition shown in~\cite{busaryev2010tracking}: 
\begin{proposition}[Busaryev et al.~\cite{busaryev2010tracking}]\label{prop:move_st}
	Given a filtration with $n$ simplices of dimensions $p-1$, $p$, and $p+1$,    let $R = DV$ denote its associated decomposition. Then, the operation $\mathrm{MoveRight}(i,j)$ constructs a valid decomposition $R' = D'V'$ in $O((\lvert \mathbb{I} \rvert + \lvert \mathbb{J} \rvert)n)$ time, where $\mathbb{I}, \mathbb{J}$ are given by:
	$$ \lvert \mathbb{I}\rvert = \sum\limits_{l=i+1}^j \mathds{1}\left(\, v_l(i) \neq 0 \, \right), \quad \mathbb{J} = \sum\limits_{l=1}^m \mathds{1}\left(\, \mathrm{low}_R(l) \in [i,j] \text{ and } r_l(i) \neq 0 \, \right)$$
	Moreover, the quantity $\lvert \mathbb{I} \rvert + \lvert \mathbb{J} \rvert$ satisfies $\lvert \mathbb{I} \rvert + \lvert \mathbb{J} \rvert \leq 2(j - i)$. 
\end{proposition}
\noindent Though similar to \emph{vineyards}, move operations confer additional advantages:
\\	
 \begin{enumerate}[labelsep=5pt, topsep=0pt,itemsep=-0.25ex,parsep=1.2ex]
 	\item[M1:] Querying the non-zero status of entries in $R$ or $V$ occurs once per move.
 	\item[M2:] $R = D V$ is \underline{not} guaranteed to be valid during the movement of $\sigma_i $ to $ \sigma_j$.
 	\item[M3:] At most $O(m)$ moves are needed to reindex $f \mapsto f'$ 
 \end{enumerate} 
 \vspace*{1em}
First, consider property M1. Prior to applying any permutation $P$ to the decomposition, it is necessary to remove non-zero entries in $V$ which render $P^TVP$ non-upper triangular, to maintain invariant I2. 
Using \emph{vineyards}, one must consistently perform $\lvert i - j \rvert - 1$ non-zero status queries interleaved between repairing column operations. A move operation, on the other hand, groups these status queries into a single pass prior to performing any modifying operations.

@@

As shown by example 2.3, \emph{moves} can be cheaper than \emph{vineyards} in terms of column operations. However, it is not immediately clear that this is always the case upon inspection of~\ref{alg:mr}, as the usage of a donor column seemingly implies that many $O(m)$ copy operations need to be performed. 
It turns out that we may handle all such operations except the first in $O(1)$ time, which we formalize below.
\begin{proposition}\label{prop:factor2}
	Let $(K,f)$ denote a filtration of size $\lvert K \rvert = m$ with decomposition $R = D V$ and let $T$ denote the number of column operations needed by the vineyards algorithm to perform the sequence of transpositions:
	$$ R = R_1 \overset{s_1}{\to} R_2 \overset{s_2}{\to} \dots \overset{s_k}{\to} R_{k-1} = R' $$
	where $s_i$ denotes the transposition $(i,i+1)$, $i < j$, and $k = \lvert i - j\rvert$. Moreover, let $M$ denote the number of column operations to perform the same update $R \mapsto R'$ with $\mathrm{Move}(i,j)$. Then the inequality $M \leq T$ holds.
\end{proposition}
\begin{proof} \normalsize
First, consider executing the \emph{vineyards} algorithm with a given pair $(i, j)$. As there are at most $2$ column operations, any contiguous sequence of transpositions $(i,i+1), (i+1,i+2), \ldots (j-1,j)$ induces at most $2(\lvert i - j \rvert)$ column operations in both $R$ and $V$, giving a total of $4(\lvert i - j \rvert)$ column operations.

Now consider a single MoveRight($i$,$j$) outlined in Algorithm~\ref{alg:mr}. Here, the dominant cost again are the column operations (line 5). 
Though we need an extra $O(m)$ storage allocation for the donor columns $d_\ast$ prior to the movement, notice that assignment to and from $d_\ast$ (lines (4), and (7) in RestoreRight of MoveRight, respectively) requires just $O(1)$ time via a pointer swapping argument. That is, when $d_\mathrm{low}' < d_\mathrm{low}$, instead of copying $\mathrm{col}_\ast(k)$ to $d_\ast'$---which takes $O(m)$ time---we instead swap their column pointers in $O(1)$   prior to column operations. After the movement, $d_\ast$ contains the newly modified column and $\mathrm{col}_\ast(k)$ contains the unmodified donor $d_\ast'$, so the final donor swap also requires $O(1)$ time. 
Since at most one $O(m)$ column operation is required for each index in $[i, j]$, moving a column from $i$ to $j$ where $i < j$ requires at most $2(\lvert i - j \rvert)$ column operations for both $R$ and $V$. The claimed inequality follows. 
%since although permutations take $O(\lvert i - j\rvert)$ time we require only $1$ such permutation per movement. 
	%Identifying non-zero entries in the matrix occurs once, which takes $O(\lvert i - j \rvert \log(m))$ time.
\end{proof}

As a final remark, we note that the combination of \emph{MoveRight} and \emph{MoveLeft} enable efficient simplex additions or deletions to the underlying complex. In particular, given $K$ and a decomposition $R = DV$, obtaining a valid decomposition $R' = D' V'$ of $K' = K \cup \{\sigma\}$ can be achieved by appending its requisite elementary chains to $D$ and $V$, reducing them, and then executing \emph{MoveLeft}$(m+1, i)$, where $i = f'(\sigma)$. Dually, deleting a simplex $\sigma_i$ may be achieved via \emph{MoveRight} by moving $i$-th to the end of the decomposition and dropping the corresponding columns. 
% with a logarithmic loss in efficiency, although executing the movement has a higher constant factor. Nonetheless, we can still move a simplex arbitrarily within the given filtration with such a procedure,  which when paired with the augmented sparse matrix structure from~\cite{} implies one can move simplices within the filtration with just O(|i - j|) time permutation, as opposed the |i - j| separate O(1) permutations. Moreover, the ability to move a simplex earlier in the filtration enables a key preprocessing step which we introduce next which, as we shall prove, may greatly reduce the number of times the decomposition need be maintained.

@@

%\subsection{Proposed approach}
We begin with a brief overview of the pipeline to compute the persistence diagrams of a discrete $1$-parameter family $(f_1, f_2, \dots, f_n)$ of filtrations. 
%$f_i : K \rightarrow [m]$ of a simplicial complex $K$. 
Without loss of generality, assume each filtration $f_i : K \rightarrow [m]$ is a simplexwise filtration of a fixed simplicial complex $K$ with $\lvert K \rvert =m$, and let $(f, f')$ denote any pair of such filtrations.  
%or equivalently, permutations of the index set $[m]$. For each $\rho_i$ we let  .
%a discrete $1$-parameter family $\mathcal{F} = (f_1, f_2, \dots, f_n)$ of filtrations $f_i : K \rightarrow [m]$ of a simplicial complex $K$ with $\lvert K \rvert =m$, and the goal being to compute the  persistence diagrams of each $(K,f_i)$.
Our strategy to efficiently obtain a valid $RV$ decomposition of the filtration $(K,f')$ from a given decomposition of $(K, f)$ is to decompose a fixed bijection $\rho : f \to f'$ satisfying $f' = \rho \circ f$ into a \emph{schedule} of updates: 
\begin{definition}[Schedule]\label{def:schedule}
Given a pair of filtrations $(K, f), (K,f')$ 
%and a fixed bijection $\rho : f \to f'$, let $\rho^\ast : [m] \to [m]$ denote its induced bijection on $S_m$ 
and $R = DV$ the initial decomposition of $(K, f)$, a \emph{schedule}  $\mathcal{S} = (s_1, s_2, \dots, s_d )$ is a sequence of permutations  satisfying: 
	\begin{equation}\label{eq:rv_seq_schedule}
   		R = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R'
   	\end{equation}
   	where, for each $i \in [d]$, $R_i = D_i V_i$ is a valid decomposition respecting invariants~\ref{inv:decomposition}, and $R'$ is a valid decomposition for $(K,f')$.
\end{definition}
\noindent
To produce this sequence of permutations $\mathcal{S} = (s_1, \dots, s_d)$ from $\rho$, our approach is as follows: let $\rho^\ast : [m] \to [m]$ denote its \emph{induced}
%\footnote{Recall that an induced representation...} 
bijection on $S_m$, and define $q = (\rho^\ast(1), \rho^\ast(2), \ldots, \rho^\ast(m) )$. 
We compute a longest increasing subsequence $\mathrm{LIS}(q)$, and then use this subsequence to recover a longest common subsequence (LCS) between $f$ and $f'$, which we denote later with $\mathrm{LCS}(f, f')$. 
We pass $q$, $\mathrm{LIS}(q)$, and a ``greedy'' heuristic for picking moves $h$ to our scheduling algorithm (to be defined later), which returns as output an ordered set of move permutations  $\mathcal{S}$ of minimum size satisfying~\eqref{eq:rv_seq_schedule}---a \emph{schedule} for the pair $(f, f')$.
These sequence of steps is outlined in Algorithm~\ref{alg:schedule} below.
%Note one need not explicitly store the family of filtrations nor the schedules between them---Algorithm~\ref{alg:schedule} may be easily modified to be completely \emph{online}, keeping at most two filtrations and one decomposition in memory at any given time. 
% Not all valid schedules are created equal: some incur a much larger number of column operations, even if they have a minimal number of moves; we address this issue in Section~\ref{sec:schedule_cost}.
% Given the set of schedules for each pair $(K_i, K_{i+1})$, we proceed to computing an $R=DV$ decomposition for the first filtration $K_1$, and then execute the moves stored in $\mathcal{S}$ in sequence. During execution, we obtain a valid decomposition $R = DV$ for each filtration $K_1, K_2, \dots, K_n$ from which we extract persistence diagrams along the way.

\begin{algorithm}[h]
	\caption{Scheduling algorithm}\label{alg:schedule}
	\setstretch{1.05}
    \begin{algorithmic}[1]
    	\Require Filtration pair $\mathcal{F} = \{f_0, f_1\}$ and valid $R = DV$ for $(K,f_0)$
    	\Ensure Valid $R = D V$ is computed for $(K, f_1)$
    	\Procedure{MoveSchedule}{$\mathcal{F} = ( f_0, f_1 )$, $R$, $V$} 
    		\State Fix $\rho \in S_m$ such that $f_1 = \rho \circ f_0$ \Comment{$O(m)$} 
    		\State $q \gets  (\rho^\ast(1), \rho^\ast(2), \ldots, \rho^\ast(m) )$ \Comment{$O(m)$}
    		\State $\mathrm{lis}_{q} \gets \mathrm{LIS}(q)$ \Comment{$O(m \log \log m)$}
%    		\State $\bar{p} \gets q^{-1} \circ p$
%    		\State $\mathcal{L} \gets \mathrm{LCS}(p,q) = \mathrm{LIS}(\bar{p})$ \Comment{$O(m\log \log m)$}
			\State $h \gets \texttt{greedy}$ \Comment{Heuristic for picking moves}
    		\State $\mathcal{S} \gets \mathrm{Schedule}(q, \mathrm{lis}_{q}, h)$ \Comment{$O(d^2\log m)$}
%	    	\State $(R, V) \gets$ \Call{Reduction}{$D = \partial (K,f_1)$}
	    	\For{$(i,j)$\textbf{ in }$\mathcal{S}$} \Comment{$\lvert \mathcal{S} \rvert = d$}
	    		\State $(R, V) \gets$ \textbf{if} $i < j$ \Call{MoveRight}{i, j} \textbf{else} \Call{MoveLeft}{i, j}
	    	\EndFor
    	\EndProcedure
	\end{algorithmic}
\end{algorithm}
%\noindent Algorithm~\ref{alg:schedule} is purely illustrative at this point, and is meant to serve as a guidepost before discussing its  components in depth. Specifically, lines (4-5) are discussed next in section~\ref{sec:schedule_sizes}, and the algorithm for line (6) is given in section~\ref{sec:schedule_cost}. Note that the loop starting at line (3) in Algorithm~\ref{alg:schedule} could always be merged with the loop starting at line (8), i.e. the individual schedules $S_i$ may be computed on the fly.

@@

\begin{proof}

@@

We conclude this section with the main theorem of this effort: an output-sensitive bound on the computation of persistence dynamically.
  \begin{theorem}\label{prop:sparsity_result}
   Given a pair of filtrations $(K,f), (K,f')$ and a decomposition $R = DV$ of $K$, the size of a minimal sequence $\mathcal{S} = \left( s_1, s_2, \dots, s_d \right)$ of cyclic `move' permutations $s_k = (i_k, j_k)$ satisfying: 
   	\begin{equation}\label{eq:rv_seq_theorem}
   		R = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R'
   	\end{equation}
   	can be determined in $O(m \log \log m)$ time and $O(m)$ space, where $R' = D_d V_d$ denotes a valid decomposition of $(K, f')$. 
   	Moreover, if $i_k < j_k$ for all $k \in [d]$, then computing~\eqref{eq:rv_seq_theorem} requires $O(\\nu)$ column operations, where:	  
   	$$  \quad \nu = \sum\limits_{k = 1}^d \, (\lvert \mathbb{I}_{k} \rvert + \lvert \mathbb{J}_{k} \rvert) $$
%	$$  \quad \kappa = \sum\limits_{k = 1}^d \, \lvert \mathbb{I}_{k}\rvert + \lvert \mathbb{J}_{k}\rvert $$
	where the quantities $\lvert \mathbb{I}_{k} \rvert, \lvert \mathbb{J}_{k} \rvert$ of the intermediate $R_k, V_k$ are given in Proposition~\ref{prop:move_st}. Note that since $\nu$ depends on the sparsity of the intermediate entries $V_1, V_2, \dots, V_d$ and $R_1, R_2, \dots, R_d$, the bound $O(\nu)$ is output-sensitive. 
%	and where $\lvert \mathbb{I}_{k} \rvert$ denotes the number of non-zero entries $V_k[i_k, i_k:j_k]$, $\lvert \mathbb{J}_{k}\rvert$ the number of columns satisfying $\mathrm{low}_{R_k} \in [i_k:j_k]$ and $\mathrm{row}_{R_k}(i_k) \neq 0$, and $d = m - \lvert \mathrm{LCS}(K, K') \rvert$.
\end{theorem}
\begin{proof} \normalsize
	Proposition~\ref{prop:schedule_size} yields the necessary conditions for constructing $\mathcal{S}$  with optimal size $d$ in $O(m \log \log m)$ time and $O(m)$. The definition of $\nu$ follows directly from Algorithm~\ref{alg:mr}. 
\end{proof}

@@

We seek a \emph{proxy objective} that correlates with~\eqref{eq:schedule_cost} and does not explicitly depend on the entries in the decomposition, as minimizing these terms directly is difficult due to  the changing sparsity of the intermediate matrices $R_k, V_k$,  
Given a pair of filtrations $(K, f), (K,f')$, consider a schedule $\mathcal{S} \in \Phi(f,f',d)$ of cyclic permutations $(i_1, j_1), (i_2, j_2), \dots, (i_d, j_d)$ minimizing:
\begin{equation}\label{eq:upper_bound_move_cost}
\mathrm{\widetilde{cost}}(\mathcal{S}) = \sum\limits_{k=1}^d 2 \lvert i_k - j_k \rvert \geq \sum\limits_{k=1}^d ( \lvert \mathbb{I}_k \rvert + \lvert \mathbb{J}_k\rvert )
In practice, this bound is very loose due to the sparsity of both $R$ and $V$. Nonetheless, the complexities of the move operations discussed in Section~\ref{sec:moves} depend on $\lvert i - j \rvert$, and minimizing~\eqref{eq:upper_bound_move_cost} has the intuitive interpretation as minimizing \emph{net displacement}. 
%The Kendall-$\tau$ distance yields an upper bound on the number of column reductions needed to perform any schedule. 
A similar $\ell_1$-type distance for measuring the disarrangement between permutations is the \emph{Spearman footrule distance}, defined as:
\begin{equation}\label{eq:spearman_dist}
F(p, q) = \sum\limits_{i =1}^m \lvert p(i) - q(i) \rvert =  \sum\limits_{i =1}^m \lvert i - (q^{-1} \circ p)(i) \rvert
The Spearman distance shares any similarities with $K_\tau$; like $K_\tau$, it is also metric on $S_m$ that is invariant under consistent relabeling. 
%and the distance between pairs of uniformly random permutations $p, q \sim S_m$ are asymptotically normal. 
Indeed, Diaconis~\cite{diaconis1977spearman} showed the Spearman distance approximates $K_\tau$ within a factor of two: 
\begin{equation}\label{eq:diaconis_inequality}
K_\tau(p, q) \leq F(p, q) \leq 2 K_\tau(p, q)
Moreover, in contrast to $K_\tau$, the Spearman distance can be computed in $O(m)$ time and is often used in the well-known \emph{rank aggregation problem}. Indeed, whereas obtaining a Kemeni optimal aggregation with respect to $K_\tau$ is NP-hard, the optimal such aggregation with respect to $F$ is obtainable in polynomial time~\cite{dinu2006efficient}. 
% Indeed, it is used...
%, and 
%For example, one of the tasks of the \emph{rank aggregation} problem is to obtain an  aggregation~\cite{} $\sigma \in S_m$ from a set of $k$ rankings $T = \{\tau_1, \tau_2, \dots, \tau_k\}$ that obeys some optimality criteria. A Kemeny optimal ranking is a permutation $\sigma^\ast$ that minimizes the sum of Kendall distances to the gievn set of rankings: 
%$$ \argmin_{\sigma \in S_m} = \sum\limits_{i = 1}^k K_\tau(\sigma, \tau_i)$$ 
%Computationally, obtaining a Kemeny optimal ranking is NP-hard\cite{}. In contrast, a 2-approximation may be obtained by replacing the Kendall distance with the Spearman distance, as in this case the problem reduces to finding a perfect matching in a bipartite graph, which can be accomplished in polynomial time. 
To adapt $F(\cdot, \cdot)$ to sortings, we decompose the Spearman distance additively via the bound:

>>>>

\begin{figure}[!htb]
    \centering
    \includegraphics[height=1.25in]{circle_complex.png}
    \caption{Top: A video of an expanding annulus. Bottom: Sublevel-set filtrations, via (negative) pixel intensity, of a Freudenthal triangulation of the plane.}
    \label{fig:vidExample}
\end{figure}

@@

%\begin{remark}
%	Conceptually, there is a conceptual similarity between the cyclic permutation that moves a simplex in a filtration the \emph{cycle sort}~\cite{haddon1990cycle}, a simple sorting algorithm method that has $O(n^2)$ time complexity for general inputs and $O(n)$ complexity for permutations.
%	Indeed, just as cycle sort rotates elements within a \emph{cycle} (in the permutation sense) using $O(1)$ auxiliary storage, the move right operation  
%\end{remark}
Regarding the complexity of move operations, which clearly depend on the sparsity of $R$ and $V$, we recall the proposition shown in~\cite{busaryev2010tracking}: 
\begin{proposition}[Busaryev et al.~\cite{busaryev2010tracking}]\label{prop:move_st}
	Given a filtration with $n$ simplices of dimensions $p-1$, $p$, and $p+1$,    let $R = DV$ denote its associated decomposition. Then, the operation $\mathrm{MoveRight}(i,j)$ constructs a valid decomposition $R' = D'V'$ in $O((\lvert \mathbb{I} \rvert + \lvert \mathbb{J} \rvert)n)$ time, where $\mathbb{I}, \mathbb{J}$ are given by:
	$$ \lvert \mathbb{I}\rvert = \sum\limits_{l=i+1}^j \mathds{1}\left(\, v_l(i) \neq 0 \, \right), \quad \mathbb{J} = \sum\limits_{l=1}^m \mathds{1}\left(\, \mathrm{low}_R(l) \in [i,j] \text{ and } r_l(i) \neq 0 \, \right)$$
	Moreover, the quantity $\lvert \mathbb{I} \rvert + \lvert \mathbb{J} \rvert$ satisfies $\lvert \mathbb{I} \rvert + \lvert \mathbb{J} \rvert \leq 2(j - i)$. 
\end{proposition}
\noindent Though similar to \emph{vineyards}, move operations confer additional advantages:
\\	
 \begin{enumerate}[labelsep=5pt, topsep=0pt,itemsep=-0.25ex,parsep=1.2ex]
 	\item[M1:] Querying the non-zero status of entries in $R$ or $V$ occurs once per move.
 	\item[M2:] $R = D V$ is \underline{not} guaranteed to be valid during the movement of $\sigma_i $ to $ \sigma_j$.
 	\item[M3:] At most $O(m)$ moves are needed to reindex $f \mapsto f'$ 
 \end{enumerate} 
 \vspace*{1em}
First, consider property M1. Prior to applying any permutation $P$ to the decomposition, it is necessary to remove non-zero entries in $V$ which render $P^TVP$ non-upper triangular, to maintain invariant I2. 
Using \emph{vineyards}, one must consistently perform $\lvert i - j \rvert - 1$ non-zero status queries interleaved between repairing column operations. A move operation groups these status queries into a single pass prior to performing any modifying operations.

@@

As shown by example 2.3, \emph{moves} can be cheaper than \emph{vineyards} in terms of column operations. However, it is not clear that this is always the case upon inspection of~\ref{alg:mr}, as the usage of a donor column seemingly implies that many $O(m)$ copy operations need to be performed. 
It turns out that we may handle all such operations except the first in $O(1)$ time, which we formalize below.
\begin{proposition}\label{prop:factor2}
	Let $(K,f)$ denote a filtration of size $\lvert K \rvert = m$ with decomposition $R = D V$ and let $T$ denote the number of column operations needed by the vineyards algorithm to perform the sequence of transpositions:
	$$ R = R_1 \overset{s_1}{\to} R_2 \overset{s_2}{\to} \dots \overset{s_k}{\to} R_{k-1} = R' $$
	where $s_i$ denotes the transposition $(i,i+1)$, $i < j$, and $k = \lvert i - j\rvert$. Moreover, let $M$ denote the number of column operations to perform the same update $R \mapsto R'$ with $\mathrm{Move}(i,j)$. Then the inequality $M \leq T$ holds.
\end{proposition}
\begin{proof} \normalsize
First, consider executing the \emph{vineyards} algorithm with a given pair $(i, j)$. As there are at most $2$ column operations, any contiguous sequence of transpositions $(i,i+1), (i+1,i+2), \ldots (j-1,j)$ induces at most $2(\lvert i - j \rvert)$ column operations in both $R$ and $V$, giving a total of $4(\lvert i - j \rvert)$ column operations.

Now consider a single MoveRight($i$,$j$) outlined in Algorithm~\ref{alg:mr}. Here, the dominant cost again are the column operations (line 5). 
Though we need an extra $O(m)$ storage allocation for the donor columns $d_\ast$ prior to the movement, notice that assignment to and from $d_\ast$ (lines (4), and (7) in RestoreRight of MoveRight, respectively) requires just $O(1)$ time via a pointer swapping argument. That is, when $d_\mathrm{low}' < d_\mathrm{low}$, instead of copying $\mathrm{col}_\ast(k)$ to $d_\ast'$---which takes $O(m)$ time---we instead swap their column pointers in $O(1)$   prior to column operations. After the movement, $d_\ast$ contains the newly modified column and $\mathrm{col}_\ast(k)$ contains the unmodified donor $d_\ast'$, so the final donor swap also requires $O(1)$ time. 
Since at most one $O(m)$ column operation is required for each index in $[i, j]$, moving a column from $i$ to $j$ where $i < j$ requires at most $2(\lvert i - j \rvert)$ column operations for both $R$ and $V$. The claimed inequality follows. 
%since although permutations take $O(\lvert i - j\rvert)$ time we require only $1$ such permutation per movement. 
	%Identifying non-zero entries in the matrix occurs once, which takes $O(\lvert i - j \rvert \log(m))$ time.
\end{proof}
As a final remark, we note that the combination of \emph{MoveRight} and \emph{MoveLeft} enable efficient simplex additions or deletions to the underlying complex. In particular, given $K$ and a decomposition $R = DV$, obtaining a valid decomposition $R' = D' V'$ of $K' = K \cup \{\sigma\}$ can be achieved by appending its requisite elementary chains to $D$ and $V$, reducing them, and then executing \emph{MoveLeft}$(m+1, i)$ with $i = f'(\sigma)$. Dually, deleting a simplex $\sigma_i$ may be achieved via \emph{MoveRight} by moving $i$-th to the end of the decomposition and dropping the corresponding columns. 
% with a logarithmic loss in efficiency, although executing the movement has a higher constant factor. Nonetheless, we can still move a simplex arbitrarily within the given filtration with such a procedure,  which when paired with the augmented sparse matrix structure from~\cite{} implies one can move simplices within the filtration with just O(|i - j|) time permutation, as opposed the |i - j| separate O(1) permutations. Moreover, the ability to move a simplex earlier in the filtration enables a key preprocessing step which we introduce next which, as we shall prove, may greatly reduce the number of times the decomposition need be maintained.

@@

%\subsection{Proposed approach}
We begin with a brief overview of the pipeline to compute the persistence diagrams of a discrete $1$-parameter family $(f_1, f_2, \dots, f_n)$ of filtrations. 
%$f_i : K \rightarrow [m]$ of a simplicial complex $K$. 
Without loss of generality, assume each filtration $f_i : K \rightarrow [m]$ is a simplexwise filtration of a fixed simplicial complex $K$ with $\lvert K \rvert =m$, and let $(f, f')$ denote any pair of such filtrations.  
%or equivalently, permutations of the index set $[m]$. For each $\rho_i$ we let  .
%a discrete $1$-parameter family $\mathcal{F} = (f_1, f_2, \dots, f_n)$ of filtrations $f_i : K \rightarrow [m]$ of a simplicial complex $K$ with $\lvert K \rvert =m$, and the goal being to compute the  persistence diagrams of each $(K,f_i)$.
Our strategy to efficiently obtain a valid $RV$ decomposition of the filtration $(K,f')$ from a given decomposition of $(K, f)$ is to decompose a fixed bijection $\rho : [m] \to [m]$ satisfying $f' = \rho \circ f$ into a \emph{schedule} of updates: 
\begin{definition}[Schedule]\label{def:schedule}
Given a pair of filtrations $(K, f), (K,f')$ 
%and a fixed bijection $\rho : f \to f'$, let $\rho^\ast : [m] \to [m]$ denote its induced bijection on $S_m$ 
and $R = DV$ the initial decomposition of $(K, f)$, a \emph{schedule}  $\mathcal{S} = (s_1, s_2, \dots, s_d )$ is a sequence of permutations  satisfying: 
	\begin{equation}\label{eq:rv_seq_schedule}
   		R = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R'
   	\end{equation}
   	where, for each $i \in [d]$, $R_i = D_i V_i$ is a valid decomposition respecting invariants~\ref{inv:decomposition}, and $R'$ is a valid decomposition for $(K,f')$.
\end{definition}
\noindent
To produce this sequence of permutations $\mathcal{S} = (s_1, \dots, s_d)$ from $\rho$, our approach is as follows:
%\footnote{Recall that an induced representation...} 
define $q = (\rho(1), \rho(2), \ldots, \rho(m) )$. 
We compute a longest increasing subsequence $\mathrm{LIS}(q)$, and then use this subsequence to recover a longest common subsequence (LCS) between $f$ and $f'$, which we denote later with $\mathrm{LCS}(f, f')$. 
We pass $q$, $\mathrm{LIS}(q)$, and a ``greedy'' heuristic for picking moves $h$ to our scheduling algorithm (to be defined later), which returns as output an ordered set of move permutations  $\mathcal{S}$ of minimum size satisfying~\eqref{eq:rv_seq_schedule}---a \emph{schedule} for the pair $(f, f')$.
These sequence of steps is outlined in Algorithm~\ref{alg:schedule} below.
%Note one need not explicitly store the family of filtrations nor the schedules between them---Algorithm~\ref{alg:schedule} may be easily modified to be completely \emph{online}, keeping at most two filtrations and one decomposition in memory at any given time. 
% Not all valid schedules are created equal: some incur a much larger number of column operations, even if they have a minimal number of moves; we address this issue in Section~\ref{sec:schedule_cost}.
% Given the set of schedules for each pair $(K_i, K_{i+1})$, we proceed to computing an $R=DV$ decomposition for the first filtration $K_1$, and then execute the moves stored in $\mathcal{S}$ in sequence. During execution, we obtain a valid decomposition $R = DV$ for each filtration $K_1, K_2, \dots, K_n$ from which we extract persistence diagrams along the way.

\begin{algorithm}[h]
	\caption{Scheduling algorithm}\label{alg:schedule}
	\setstretch{1.05}
    \begin{algorithmic}[1]
    	\Require Filtration pair $\mathcal{F} = \{f_0, f_1\}$ and valid $R = DV$ for $(K,f_0)$
    	\Ensure Valid $R = D V$ is computed for $(K, f_1)$
    	\Procedure{MoveSchedule}{$\mathcal{F} = ( f_0, f_1 )$, $R$, $V$} 
    		\State Fix $\rho \in S_m$ such that $f_1 = \rho \circ f_0$ \Comment{$O(m)$} 
    		\State $q \gets  (\rho(1), \rho(2), \ldots, \rho(m) )$ \Comment{$O(m)$}
    		\State $\mathrm{lis}_{q} \gets \mathrm{LIS}(q)$ \Comment{$O(m \log \log m)$}
%    		\State $\bar{p} \gets q^{-1} \circ p$
%    		\State $\mathcal{L} \gets \mathrm{LCS}(p,q) = \mathrm{LIS}(\bar{p})$ \Comment{$O(m\log \log m)$}
			\State $h \gets \texttt{greedy}$ \Comment{Heuristic for picking moves}
    		\State $\mathcal{S} \gets \mathrm{Schedule}(q, \mathrm{lis}_{q}, h)$ \Comment{$O(d^2\log m)$}
%	    	\State $(R, V) \gets$ \Call{Reduction}{$D = \partial (K,f_1)$}
	    	\For{$(i,j)$\textbf{ in }$\mathcal{S}$} \Comment{$\lvert \mathcal{S} \rvert = d$}
	    		\State $(R, V) \gets$ \textbf{if} $i < j$ \Call{MoveRight}{i, j} \textbf{else} \Call{MoveLeft}{i, j}
	    	\EndFor
    	\EndProcedure
	\end{algorithmic}
\end{algorithm}
%\noindent Algorithm~\ref{alg:schedule} is purely illustrative at this point, and is meant to serve as a guidepost before discussing its  components in depth. Specifically, lines (4-5) are discussed next in section~\ref{sec:schedule_sizes}, and the algorithm for line (6) is given in section~\ref{sec:schedule_cost}. Note that the loop starting at line (3) in Algorithm~\ref{alg:schedule} could always be merged with the loop starting at line (8), i.e. the individual schedules $S_i$ may be computed on the fly.

@@

@@

We conclude this section with the main theorem of this effort: an output-sensitive bound on the computation of persistence dynamically.
  \begin{theorem}\label{prop:sparsity_result}
   Given a pair of filtrations $(K,f), (K,f')$, a decomposition $R = DV$ of $K$, and a sequence $\mathcal{S} = \left( s_1, s_2, \dots, s_d \right)$ of cyclic `move' permutations $s_k = (i_k, j_k)$ satisfying $i_k < j_k$ for all $k \in [d]$, computing the updates:
   \begin{equation}\label{eq:rv_seq_theorem}
   		R = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R'
   	\end{equation}
   	where $R' = D_d V_d$ denotes a valid decomposition of $(K, f')$ requires $O(\nu)$ column operations, where $\nu$ depends on the sparsity of the intermediate entries $V_1, V_2, \dots, V_d$ and $R_1, R_2, \dots, R_d$:	  
   	$$  \quad \nu = \sum\limits_{k = 1}^d \, (\lvert \mathbb{I}_{k} \rvert + \lvert \mathbb{J}_{k} \rvert), \quad \text{ where }  \lvert \mathbb{I}_{k} \rvert, \lvert \mathbb{J}_{k} \rvert \text{are given in Proposition~\ref{prop:move_st} } $$ 
   	Moreover, the size of a minimal such $\mathcal{S}$ can be determined in $O(m \log \log m)$ time and $O(m)$ space. 
%	$$  \quad \kappa = \sum\limits_{k = 1}^d \, \lvert \mathbb{I}_{k}\rvert + \lvert \mathbb{J}_{k}\rvert $$
	%	and where $\lvert \mathbb{I}_{k} \rvert$ denotes the number of non-zero entries $V_k[i_k, i_k:j_k]$, $\lvert \mathbb{J}_{k}\rvert$ the number of columns satisfying $\mathrm{low}_{R_k} \in [i_k:j_k]$ and $\mathrm{row}_{R_k}(i_k) \neq 0$, and $d = m - \lvert \mathrm{LCS}(K, K') \rvert$.
\end{theorem}
\begin{proof} \normalsize
	Proposition~\ref{prop:schedule_size} yields the necessary conditions for constructing $\mathcal{S}$  with optimal size $d$ in $O(m \log \log m)$ time and $O(m)$. The definition of $\nu$ follows directly from Algorithm~\ref{alg:mr}. 
\end{proof}

@@

In light of section~\ref{sec:greedy}, we seek an alternative objective that correlates with~\eqref{eq:schedule_cost} and does not depend on the entries in the decomposition.
Given a pair of filtrations $(K, f), (K,f')$, a natural schedule $\mathcal{S} \in \Phi(f,f',d)$ of cyclic permutations $(i_1, j_1), (i_2, j_2), \dots, (i_d, j_d)$ is one minimizing the upper bound:
\begin{equation}\label{eq:upper_bound_move_cost}
\mathrm{\widetilde{cost}}(\mathcal{S}) = \sum\limits_{k=1}^d 2 \lvert i_k - j_k \rvert \geq \sum\limits_{k=1}^d ( \lvert \mathbb{I}_k \rvert + \lvert \mathbb{J}_k\rvert )
\end{equation}
%In practice, this bound is very loose due to the sparsity of both $R$ and $V$. Nonetheless, the objective is intuitive, and the complexities of the move operations discussed in Section~\ref{sec:moves} depend on $\lvert i - j \rvert$. 
%and minimizing~\eqref{eq:upper_bound_move_cost} has the intuitive interpretation as minimizing \emph{net displacement}. 
Unfortunately, even obtaining an optimal schedule $\mathcal{S}^\ast \in \Phi(f,f',d)$ minimizing~\eqref{eq:upper_bound_move_cost}
% $\mathcal{S}^\ast = \argmin_{\mathcal{S} \in \Phi(f,f',d)} \mathrm{\widetilde{cost}}(\mathcal{S})$
does not appear tractable due to its similarity with the $k$-layer crossing minimization problem, which is NP-hard for $k$ sets of permutations when $k \geq 4$~\cite{biedl2009complexity}. For additional discussion on the relationship between these two problems, see section~\ref{sec:cross_minimization} in the appendix.
